The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-29 paper.tex
Introducao

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms.
%
Employing the synchronous programming model, it allows for a simple reasoning 
about shared-memory concurrency.
Furthermore, its restricted semantics enables deterministic and memory-safe 
programs.
%
% TODO: some uses?
%
In this work, we propose a formal semantics for \CEU focusing on its particular 
control mechanisms, such as parallel compositions, finalization, and 
stack-based internal events.
%
We also present an implementation with two backends:
one aiming for resource efficiency and interoperability with C, and another for 
code dissemination in sensor networks.
\end{abstract}

@@

\begin{itemize}
\item Relevancia de Esterel
\item Ceu
    \begin{itemize}
        \item Foco inicial em "constrained WSNs"
        \item Estendendo para outros dominios? (jogos, artigo Mod'15)
        \item Usos: SenSys, Terra, sala de aula, GSoC
        \item ideia de tamanhos, ROM,RAM,CPU
    \end{itemize}
\item Limitacoes do modelo sincrono
\item Semantica e Implementacao (nesse artigo, somente subset estatico)
% Artigo de Rust, nao resolve os problemas
%https://blog.skcript.com/asynchronous-io-in-rust-36b623e7b965
%state machines everywhere
\end{itemize}

@@

\section{Formal Semantics}

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the main differences to Esterel, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and $C$ calls (which behave like in conventional 
imperative languages).

@@

The $mem(id)$ primitive represents all accesses, assignments, and $C$ function 
calls that affect a memory location identified by $id$.
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
%considered to be atomic and instantaneous
%
The special notation $nop$ is used to represent an innocuous $mem$ expression 
(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

\subsection{Operational Semantics}
\label{sec.sem}

@@

\section{Implementation}

@@

This phase detects inconsistencies in \CEU programs, such as unbounded loops, 
suspicious accesses to shared memory, and also ``classical'' semantic analysis, 
such as building a symbol table for checking variable declarations.
%
This phase outputs code in $C$, given the way \CEU is tied to $C$ by design.
%
Some type checking is delayed to the last phase to take advantage of 
\code{gcc}'s error handling.
Therefore, we annotate the $C$ output file with \code{\#line} pragmas matching 
the original file in \CEU.

@@

The final phase packs the generated $C$ file with the \CEU runtime and 
platform-dependent functionality, compiling them with \emph{gcc} and generating 
the final binary.
%
The \CEU runtime comprehends the scheduler, timer management, and the external 
$C$ API.
%
The platform files include libraries for I/O and bindings to invoke the \CEU 
scheduler on external events.

@@

A node inherits the set $I$ from its direct parent and calculates $O$ according 
to its type:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, $C$ calls, and 
declarations simply reproduce $O=I$, as they do not await;
%
\item An \code{await e} statement has $O=\{e\}$.
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never terminates, i.e., never proceeds to the statement 
immediately following it (see also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$.
The first child inherits $I$ from its parent node, and the set $O$ for the 
sequence node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
O_{body}$), as the loop is also reached from its own body.
The union of all \code{break} statements' $O$ forms the set $O$ for a 
\code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$.
%
\item A parallel composition may terminate from any of its branches, hence $O = 
O_1 \cup ... \cup O_n$.
\end{itemize}

@@

\CEU favors a fine-grained use of trails, being common the use of trails that 
await a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to $C$ is straightforward~\cite{wsn.osm,wsn.ocram}: 
memory for blocks in sequence are packed in a \code{struct}, while blocks in 
parallel, in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

@@

\subsection{The External $C$ API}

As a reactive language, the execution of a program in \CEU is guided entirely 
by the occurrence of external events.
From the implementation perspective, there are three external sources of input 
into programs, which are all exposed as functions in a $C$ API:

@@

\begin{comment}
Note that \CEU{} code running from a call to \code{ceu\_go\_async} may emit an 
input event or the passage of time.
In this case, the $C$ implementation makes a tail call to the corresponding 
handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as synchronous 
code has higher priority.

@@

\section{Terra and the \CEU-VM}

\section{Related Work}

@@

A number of synchronous languages have been designed to interoperate with C, 
such as \emph{Reactive C}~\cite{rp.rc}, Protothreads \cite{wsn.protothreads}, 
\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
%
They offer Esterel-like parallel compositions with communication via shared 
variables, relying on deterministic scheduling to preserve determinism.
%
However, it is the responsibility of the programmer to specify the execution 
order for threads, based on either explicit priorities, or source code lexical 
order (like \CEU).
%
These languages have a tick-based notion of time similar to Esterel, which 
prevents the event-based temporal analysis of \CEU.

@@

\section{Final Remarks}

>>>>

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms.
%
Employing the synchronous programming model, it allows for a simple reasoning 
about shared-memory concurrency.
Furthermore, its restricted semantics enables deterministic and memory-safe 
programs.
%
% TODO: some uses?
%
In this work, we propose a formal semantics for \CEU focusing on its particular 
control mechanisms, such as parallel compositions, finalization, and 
stack-based internal events.
%
We also present an implementation with two backends:
one aiming for resource efficiency and interoperability with C, and another for 
dynamic reprogramming over the air.
\end{abstract}

@@

An established alternative to \emph{C} in the field of embedded systems is the 
family of reactive synchronous languages \cite{rp.twelve}.
%
Two major styles of synchronous languages have evolved:
in the \emph{control}--\emph{imperative} style, programs are structured with 
control flow primitives, such as parallelism, repetition, and preemption;
in the \emph{dataflow}--\emph{declarative} style, programs can be seen as 
graphs of values, in which a change to a value is propagated through its 
dependencies without explicit programming.
%
Among the control-based languages, Esterel~\cite{esterel.ieee91} influenced a 
number of other embedded languages, such as \emph{Reactive-C}~\cite{rp.rc}, 
\emph{OSM}~\cite{wsn.osm}, \emph{Sync-C}~\cite{rp.synchc}, and 
\emph{PRET-C}~\cite{rp.pretc}.

\CEU is another language targeting embedded systems based on Esterel with some 
novel functionalities:
%
\begin{itemize}
%
\item Stack-based internal events, which provide a limited form of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution.
%
\item First-class synchronized timers.
\end{itemize}

\CEU shares limitations with Esterel and synchronous languages in general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
elegantly implemented.

In this work, we present a formal semantics of a small synchronous kernel that 
represents a subset of \CEU covering these new functionalities.
%
We also present an implementation of \CEU with two backends:
one aiming for efficiency and interoperability with \emph{C},
and another based on a virtual machine that allows remote reprogramming.

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of constrained wireless sensor networks, developing a number of applications, 
protocols, and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\emph{C}, and observed a reduction in source code size (around 25\%) with a 
small increase in memory usage (around 5--10\% for ROM and 
RAM)~\cite{ceu.sensys13}.
%
Considering the \emph{VM} backend, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages.
In contrast, without the preinstalled \emph{VM} with basic \emph{I/O} 
functionality, a \emph{C}-compiled version occupies more than 2 
kbytes~\cite{ceu.terra}.

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} gives an overview of \CEU, exposing its fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \emph{C} and \emph{VM} implementation 
backends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

%\begin{itemize}
        %\item Estendendo para outros dominios? (jogos, artigo Mod'15)
        %\item Usos: sala de aula, GSoC
%\item Semantica e Implementacao (nesse artigo, somente subset estatico)
% Artigo de Rust, nao resolve os problemas
%https://blog.skcript.com/asynchronous-io-in-rust-36b623e7b965
%state machines everywhere
%\end{itemize}

@@

\section{Formal Semantics}
\label{sec.sem}

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the main differences to Esterel, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and \emph{C} calls (which behave like in conventional 
imperative languages).

@@

The $mem(id)$ primitive represents all accesses, assignments, and \emph{C} 
function calls that affect a memory location identified by $id$.
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
%considered to be atomic and instantaneous
%
The special notation $nop$ is used to represent an innocuous $mem$ expression 
(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

\subsection{Operational Semantics}

@@

\section{Implementation}
\label{sec.impl}

@@

This phase detects inconsistencies in \CEU programs, such as unbounded loops, 
suspicious accesses to shared memory, and also ``classical'' semantic analysis, 
such as building a symbol table for checking variable declarations.
%
This phase outputs code in \emph{C}, given the way \CEU is tied to \emph{C} by 
design.
%
Some type checking is delayed to the last phase to take advantage of 
\code{gcc}'s error handling.
Therefore, we annotate the \emph{C} output file with \code{\#line} pragmas 
matching the original file in \CEU.

@@

The final phase packs the generated \emph{C} file with the \CEU runtime and 
platform-dependent functionality, compiling them with \emph{gcc} and generating 
the final binary.
%
The \CEU runtime comprehends the scheduler, timer management, and the external 
\emph{C} API.
%
The platform files include libraries for I/O and bindings to invoke the \CEU 
scheduler on external events.

@@

A node inherits the set $I$ from its direct parent and calculates $O$ according 
to its type:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, \emph{C} calls, and 
declarations simply reproduce $O=I$, as they do not await;
%
\item An \code{await e} statement has $O=\{e\}$.
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never terminates, i.e., never proceeds to the statement 
immediately following it (see also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$.
The first child inherits $I$ from its parent node, and the set $O$ for the 
sequence node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
O_{body}$), as the loop is also reached from its own body.
The union of all \code{break} statements' $O$ forms the set $O$ for a 
\code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$.
%
\item A parallel composition may terminate from any of its branches, hence $O = 
O_1 \cup ... \cup O_n$.
\end{itemize}

@@

\CEU favors a fine-grained use of trails, being common the use of trails that 
await a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \emph{C} is straightforward~\cite{wsn.osm,wsn.ocram}: 
memory for blocks in sequence are packed in a \code{struct}, while blocks in 
parallel, in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

@@

\subsection{The External \emph{C} API}

As a reactive language, the execution of a program in \CEU is guided entirely 
by the occurrence of external events.
From the implementation perspective, there are three external sources of input 
into programs, which are all exposed as functions in a \emph{C} API:

@@

\begin{comment}
Note that \CEU{} code running from a call to \code{ceu\_go\_async} may emit an 
input event or the passage of time.
In this case, the \emph{C} implementation makes a tail call to the 
corresponding handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as 
synchronous code has higher priority.

@@

\subsection{The Terra Virtual Machine}
\label{sec.terra}

\section{Related Work}
\label{sec.related}

@@

A number of synchronous languages have been designed to interoperate with C, 
such as \emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
%
They offer Esterel-like parallel compositions with communication via shared 
variables, relying on deterministic scheduling to preserve determinism.
%
However, it is the responsibility of the programmer to specify the execution 
order for threads, based on either explicit priorities, or source code lexical 
order (like \CEU).
%
These languages have a tick-based notion of time similar to Esterel, which 
prevents the event-based temporal analysis of \CEU.

@@

\section{Final Remarks}
\label{sec.conclusion}
